---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# rho.agent

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

[`rho.agent`](https://rgenomicsetl.github.io/Rho/rho.agent/) is Rho's
provider-neutral, multi-turn agent loop. It turns provider streams into an
ordered lifecycle, schedules typed tools, drains steering and follow-up queues,
and propagates cancellation without placing coding or bioinformatics policy in
the core.

## Run an agent

```{r agent}
library(rho.async)
library(rho.ai)
library(rho.agent)

agent <- rho_agent(
  rho_faux_provider(),
  rho_model(provider = "faux", id = "faux")
)
task <- rho_prompt(agent, "show the protocol")
rho_is_task(task)

run <- rho_await(task, timeout = 5000)
list(
  status = run@status,
  answer = run@messages[[2L]]@content[[1L]]@text,
  events = length(run@events)
)
```

Listeners are awaited in registration order. Tool calls that declare
`ToolMayOverlap` may run concurrently, while results remain in source order;
`ToolRequiresExclusiveExecution` makes stateful semantics explicit. Agent
policy is an S7 protocol, so applications can override context transforms and
before/after-tool decisions without replacing the loop.

The session ledger is append-only. `rho_compact()` adds a semantic checkpoint
while retaining the complete transcript, and `rho_build_agent_context()`
projects the messages sent to a provider. Threshold compaction and one retry for
a typed provider input-limit value use the same compactor and policy generics.

`rho.ext` is not yet bound to this lifecycle; the integration-or-removal proof
is tracked in [issue #9](https://github.com/RGenomicsETL/Rho/issues/9). See the
[`rho.agent` reference](https://rgenomicsetl.github.io/Rho/rho.agent/reference/).
