---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

# Rho

<!-- badges: start -->
[![R-CMD-check](https://github.com/RGenomicsETL/Rho/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/RGenomicsETL/Rho/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

**An asynchronous agent runtime for R, inspired by
[Pi](https://github.com/badlogic/pi-mono).**

Rho expresses Pi's provider and agent architecture through S7 classes and open
generics. nanonext provides asynchronous I/O, and mirai provides worker
evaluation.

An effectful operation returns a task or stream. Waiting is an explicit edge
operation, so a CLI can block while a Shiny application, extension, or another
agent keeps composing work.

## An agent run

The deterministic provider is the executable specification for the same event
protocol used by live providers. `rho_prompt()` returns before the result is
collected; `rho_await()` is the explicit wait.

```{r agent}
library(rho)

agent <- rho_agent(
  provider = rho_faux_provider(),
  model = rho_model(provider = "faux", id = "faux")
)

run_task <- rho_prompt(agent, "hello from R")
rho_is_task(run_task)

run <- rho_await(run_task, timeout = 5000)
c(
  status = run@status,
  answer = run@messages[[2L]]@content[[1L]]@text
)
```

The transcript contains typed messages and content; the lifecycle is an ordered
sequence of typed events rather than callbacks hidden inside the provider.

```{r events}
vapply(run@events, function(event) event@type, character(1))
```

## OpenAI Codex with explicit credentials

Live authentication is a value passed to the provider catalog. Rho does not
search environment variables or process-global credential state. The importer
accepts a Pi or Codex auth file only when its path is supplied by the caller.

```{r codex-agent, eval = isTRUE(getOption("rho.codex_readme_rebuild", FALSE))}
credential_path <- getOption("rho.openai_codex_credential")
credential <- rho_load_openai_codex_credential(
  path = credential_path
) |>
  rho_await(timeout = 5000)

codex_provider <- rho_openai_codex_provider()
models <- rho_models(
  providers = list(codex_provider),
  credentials = rho_memory_credential_store(
    list(`openai-codex` = credential)
  )
)

codex_model <- rho_openai_codex_model("gpt-5.3-codex-spark")
codex_agent <- rho_agent(
  provider = models,
  model = codex_model,
  tools = list(rho_tool_r()),
  stream_options = list(reasoning_effort = "minimal")
)

codex_run <- rho_prompt(
  codex_agent,
  paste(
    "Call the r tool exactly once with code sum((1:100)^2).",
    "Then answer with only the integer result."
  )
) |>
  rho_await(timeout = 120000)

tool_result <- codex_run@tool_results[[1L]]
result <- tool_result@content[[1L]]@text

codex_example <- data.frame(
  model = codex_model@id,
  status = codex_run@status,
  tool = tool_result@tool_name,
  result = result
)
codex_example
```

Rebuild this example by supplying the credential file explicitly:

```bash
Rscript scripts/render-readmes.R /absolute/path/to/auth.json
```

## Package layout

`rho` is one installable R package. Its task, HTTP, provider, agent, extension,
compute, graphics, coding, and bioinformatics contracts remain distinct
components within one namespace. DuckDB, mirai, and httr2/curl backends use
optional dependencies. Bounded test assertions are internal utilities.

The components have separate responsibilities, not separate package boundaries. The current
`rho.ext` runtime is not connected to `rho.agent`; `rho.bio.agent` only lists
registered manifests; the DuckDB SQL guard is lexical; and coding file/Bash
tools still inherit host filesystem and environment authority. These gaps are
tracked in issues [#7](https://github.com/RGenomicsETL/Rho/issues/7),
[#9](https://github.com/RGenomicsETL/Rho/issues/9), and
[#11](https://github.com/RGenomicsETL/Rho/issues/11).

Provider implementations live in `rho`. OpenAI Codex, GitHub Copilot, Z.ai, OpenAI, Anthropic, Ollama, and
the deterministic faux provider therefore share one typed provider surface
without a package per API. The [Pi parity ledger](https://rgenomicsetl.github.io/Rho/docs/pi-parity.html) distinguishes
complete wire adapters from request translators whose normalized stream is not
yet complete. OpenAI, OpenAI Codex, GitHub Copilot, Z.ai, and Anthropic have
executable normalized-stream fixtures; the ledger records the executable and
external-account evidence for each adapter.

Bioinformatics contracts and adapters consume the provider and agent substrate
without defining it.

## Install and develop

Rho targets R 4.6 or newer. Install the development package from a checkout.
The required nanonext development build comes from the pinned
[RGenomicsETL fork](https://github.com/RGenomicsETL/nanonext/tree/084c8eb82a35c38b2aaf7dd203414829c2bd8011),
not the r-lib release; the `Remotes` field declares this source.

```bash
git clone git@github.com:RGenomicsETL/Rho.git
cd Rho
Rscript -e 'remotes::install_deps(dependencies = TRUE)'
make install
```

The authored API documentation is roxygen; the authored tests are R Markdown
files under `inst/tinytest/rmd/`. Generated manuals, the namespace, executable
tests, and the README are reproducible from their sources.

```bash
make format       # Air
make rd           # roxygen2
make purl-tests   # Rmd tests -> executable tinytest files
make rdm          # rebuild README.md
make test         # run unified tinytest suite
make check        # R CMD build and R CMD check --no-manual
```

The [Pi parity ledger](https://rgenomicsetl.github.io/Rho/docs/pi-parity.html) records behavioral contracts and the
fixtures that verify them. Public release and addition to the
[RGenomicsETL R-universe](https://rgenomicsetl.r-universe.dev) follow
green package checks, live provider checks, documentation, and secret scanning.
The exact sequence is recorded in the [publishing guide](https://rgenomicsetl.github.io/Rho/docs/releasing.html).

## License

MIT.
